Analysis of Cartesian Genetic Programming's Evolutionary Mechanisms

### 2013-12-02 CGP Journal/main.tex
Cut down to length, back on 15 pages!

\newcommand{\graphicthird}[1]
{\includegraphics[width=.285\textwidth,height=.285\textheight,keepaspectratio]{#1}}

@@

In Section~\ref{sec:cgp} we provide
an introduction to CGP, with sufficient background to put later sections into
context.  Section~\ref{sec:duplicate} describes previous work on preventing mutation
from creating identifiably wasted evaluations.
Section~\ref{sec:ordering} discusses the impact of CGP's genome ordering,
along with methods to both examine and bypass the limitations imposed by that ordering.
Each technique is then rigorously tuned and qualitatively compared on a collection
of benchmarks in Section~\ref{sec:quality}.  Section~\ref{sec:analysis} then
provides a deep look into the details of CGP's search, and how each method impacts
search, with final remarks given in Section~\ref{sec:conclusion}.

@@

The number of wasted evaluations is
directly related to the mutation probability and the number of active genes.  With
sufficiently high probabilities and enough active genes, the chance of
creating an offspring which is actively identical to its parent is very low.
Yet such a configuration may make it difficult for CGP to optimize, as mutation
is likely to change many active genes in each application.  As was discussed
in~\cite{goldman:2013:cgpwaste}, this can make normal CGP very sensitive to the
mutation probability.

@@

\subsection{Accumulate}
Similar to \emph{Skip}, \emph{Accumulate} works by adding a step between offspring
creation and offspring evaluation.  Instead of skipping evaluations when offspring
are determined to be actively identical to their parents, \emph{Accumulate} enters
into a cycle of repeated mutation until an individual worth evaluating is created.
To understand this technique, consider an example where the parent $P$ produces
the offspring $F_0$ using mutation. If $F_0$ is actively identical $P$, $F_0$
creates its own offspring $F_1$ using mutation.  This process of $F_i$ producing
$F_{i+1}$ continues until $F_n$ is produced, where $F_n$ is not actively identical
to $P$.  $F_n$ is then evaluated.  If $F_n$ is no worse than $P$, it replaces
$F_0$, and evolution continues as though $P$ had produced $F_n$ directly.  If
$F_n$ is worse than $P$, $F_{n-1}$ replaces $F_0$, as $F_{n-1}$ has accumulated
the most mutations to inactive genes without reducing fitness.  Note that
the mutation used each time is identical to normal CGP's mutation, such that
$F_n$ will have one or more active genes different from $P$.  Furthermore, even
though this technique produces $n+1$ individuals, only $F_n$ is evaluated.
Viewed in another way, \emph{Accumulate} performs a micro evolution on each offspring
before it is evaluated, with an emphasis on encouraging neutral drift.

@@

As an extension to the theory presented in~\cite{goldman:2013:ordering}, we have
derived the formula which predicts the probability a node is active based on its
index, given in Equation~\ref{eq:prob} and plotted in Figure~\ref{fig:prob}.
\begin{equation}
p(i) \leftarrow 1 - (\frac{N+I-1}{N+I})^O \cdot \prod_{j=i+1}^{N} 1 - p(j) \cdot (1-(\frac{j+I-1}{j+I})^A)
\label{eq:prob}
\end{equation}
In this equation, $i$ is the index of the working node, $N$ is the genome size, $I$ is the number of input locations,
$O$ is the number of output locations, and $A$ is the arity of each node.
This equation is constructed as a negation, such that we determine the probability
the node is not active due to each potential source that could make it active.
The product term calculates the probability that the node at index $j$ is connected to the
node at index $i$, multiplied by the probability that the node at index $j$ is active.  This is
negated to determine the probability node $i$ is not active because of node $j$.
All together, the product term calculates the probability that no active node is
connected to the node at index $i$.  Nodes can also become active if they are directly
connected to by an output location.  As such the product is scaled by the probability
that all output locations connect to inputs or nodes not at index $i$.  This
formula assumes only operator bias, not selective pressure, is acting on the
genome.

\begin{figure}
\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{active_prob}
\caption{Plot of Equation~\ref{eq:prob} describing the probability that a node
         at a given index will be active.  $N=2000, A=2, I=6, O=6$.
         Predicts approximately 160 total active nodes, with 80\% of the first
         100 nodes active and 4\% of the remaining 1900 nodes active.}
\label{fig:prob}
\end{figure}

From Figure~\ref{fig:prob} it is clear that nodes near the beginning of the genome
are almost certainly going to be active, while nodes in the last three quarters
of the genome are almost never going to be active.

@@

%The above paragraphs were replaced by the below paragraph.
In Section~\ref{sec:ordering} we used Equation~\ref{eq:prob} to suggest \emph{Normal}
as a strong bias on the location of active nodes regardless of fitness landscape, and that
\emph{Reorder} and \emph{DAG} would likely have less or no bias.
Figure~\ref{fig:prob} shows the expected probability given by Equation~\ref{eq:prob},
using the same genome size, arity, number of inputs, and number of outputs as our
3-bit multiply.  This curve bears a striking resemblance to all three \emph{Normal}
plots in Table~\ref{tab:never_active}.

@@

Each diagram describes node behavior as falling into eight distinct categories
split between active and inactive nodes.  Nodes are put into the first category
they satisfy.
\begin{enumerate}
\item ``Constant'' - Independent of problem input, constant nodes produce a fixed output.
\item ``Used'' - Nodes still active after applying \Call{Simplify}{}.
\item ``Useful'' - Nodes with identical semantics to a ``Used'' node.
\item ``Intron'' - Any node with semantics identical to a node that was
                   active before applying \Call{Simplify}{}, but was inactive afterward.
\item ``Explore'' - All remaining nodes.
\end{enumerate}
Used nodes are by definition only in the active portion of the genome.  Similarly
Explore nodes can only be in the inactive portion of the genome.   Note that
all never active nodes are ignored by these plots, as their behavior is meaningless.

@@

\section{Conclusion}
\label{sec:conclusion}
Even with CGP's relatively simple evolutionary mechanisms, fully understanding
the effects those mechanisms have on search takes concerted effort.  By examining
the effectiveness as well as the actual behavior of traditional CGP and
a handful of variants we have helped advance this understanding.

>>>>

\newcommand{\graphicthird}[1]
{\includegraphics[width=.27\textwidth,height=.27\textheight,keepaspectratio]{#1}}

@@

\begin{comment}
In Section~\ref{sec:cgp} we provide
an introduction to CGP, with sufficient background to put later sections into
context.  Section~\ref{sec:duplicate} describes previous work on preventing mutation
from creating identifiably wasted evaluations.
Section~\ref{sec:ordering} discusses the impact of CGP's genome ordering,
along with methods to both examine and bypass the limitations imposed by that ordering.
Each technique is then rigorously tuned and qualitatively compared on a collection
of benchmarks in Section~\ref{sec:quality}.  Section~\ref{sec:analysis} then
provides a deep look into the details of CGP's search, and how each method impacts
search, with final remarks given in Section~\ref{sec:conclusion}.
\end{comment}

@@

\begin{comment}
The number of wasted evaluations is
directly related to the mutation probability and the number of active genes.  With
sufficiently high probabilities and enough active genes, the chance of
creating an offspring which is actively identical to its parent is very low.
Yet such a configuration may make it difficult for CGP to optimize, as mutation
is likely to change many active genes in each application.  As was discussed
in~\cite{goldman:2013:cgpwaste}, this can make normal CGP very sensitive to the
mutation probability.
\end{comment}

@@

\subsection{Accumulate}
Similar to \emph{Skip}, \emph{Accumulate} works by adding a step between offspring
creation and offspring evaluation.  Instead of skipping evaluations when offspring
are determined to be actively identical to their parents, \emph{Accumulate} enters
into a cycle of repeated mutation until an individual worth evaluating is created.
To understand this technique, consider an example where the parent $P$ produces
the offspring $F_0$ using mutation. If $F_0$ is actively identical to $P$, $F_0$
creates its own offspring $F_1$ using mutation.  This process of $F_i$ producing
$F_{i+1}$ continues until $F_n$ is produced, where $F_n$ is not actively identical
to $P$.  $F_n$ is then evaluated.  If $F_n$ is no worse than $P$, it replaces
$F_0$, and evolution continues as though $P$ had produced $F_n$ directly.  If
$F_n$ is worse than $P$, $F_{n-1}$ replaces $F_0$, as $F_{n-1}$ has accumulated
the most mutations to inactive genes without reducing fitness.  Note that
the mutation used each time is identical to normal CGP's mutation, such that
$F_n$ will have one or more active genes different from $P$.  Furthermore, even
though this technique produces $n+1$ individuals, only $F_n$ is evaluated.
Viewed in another way, \emph{Accumulate} performs a micro evolution on each offspring
before it is evaluated, with an emphasis on encouraging neutral drift.

@@

As an extension to the theory presented in~\cite{goldman:2013:ordering}, we have
derived the formula which predicts the probability a node is active based on its
index, given in Equation~\ref{eq:prob}.
\begin{equation}
p(i) \leftarrow 1 - (\frac{N+I-1}{N+I})^O \cdot \prod_{j=i+1}^{N} 1 - p(j) \cdot (1-(\frac{j+I-1}{j+I})^A)
\label{eq:prob}
\end{equation}
In this equation, $i$ is the index of the working node, $N$ is the genome size, $I$ is the number of input locations,
$O$ is the number of output locations, and $A$ is the arity of each node.
This equation is constructed as a negation, such that we determine the probability
the node is not active due to each potential source that could make it active.
The product term calculates the probability that the node at index $j$ is connected to the
node at index $i$, multiplied by the probability that the node at index $j$ is active.  This is
negated to determine the probability node $i$ is not active because of node $j$.
All together, the product term calculates the probability that no active node is
connected to the node at index $i$.  Nodes can also become active if they are directly
connected to by an output location.  As such the product is scaled by the probability
that all output locations connect to inputs or nodes not at index $i$.  This
formula assumes only operator bias, not selective pressure, is acting on the
genome.

\begin{comment}
\begin{figure}
\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{active_prob}
\caption{Plot of Equation~\ref{eq:prob} describing the probability that a node
         at a given index will be active.  $N=2000, A=2, I=6, O=6$.
         Predicts approximately 160 total active nodes, with 80\% of the first
         100 nodes active and 4\% of the remaining 1900 nodes active.}
\label{fig:prob}
\end{figure}

From Figure~\ref{fig:prob} it is clear that nodes near the beginning of the genome
are almost certainly going to be active, while nodes in the last three quarters
of the genome are almost never going to be active.
\end{comment}
%NEW
Substituting $N=2000$, $A=2$, $I=6$, $O=6$ into Equation~\ref{eq:prob}
predicts only 160 active nodes, with 80\% of the first 100 nodes expected to be
active while just 4\% of the remaining 1900 nodes are expected to be active.
%/NEW

@@

%The above paragraphs were replaced by the below paragraph.
In Section~\ref{sec:ordering} we used Equation~\ref{eq:prob} to suggest \emph{Normal}
as a strong bias on the location of active nodes regardless of fitness landscape, and that
\emph{Reorder} and \emph{DAG} would likely have less or no bias.
%NEW
The example inputs used in that section match those of \emph{Normal}'s tuned
configuration on 3-bit multiply.  The predicted number of active nodes and their
location in the genome match almost exactly
the behavior displayed in all three \emph{Normal} plots in Table~\ref{tab:never_active}.
%/NEW

@@

Each diagram describes node behavior as falling into eight distinct categories
split between active and inactive nodes.  Nodes are put into the first category
they satisfy.
\\
\\
\begin{enumerate}
\item ``Constant'' - Independent of problem input, constant nodes produce a fixed output.
\item ``Used'' - Nodes still active after applying \Call{Simplify}{}.
\item ``Useful'' - Nodes with identical semantics to a ``Used'' node.
\item ``Intron'' - Any node with semantics identical to a node that was
                   active before applying \Call{Simplify}{}, but was inactive afterward.
\item ``Explore'' - All remaining nodes.
\end{enumerate}
Used nodes are by definition only in the active portion of the genome.  Similarly
Explore nodes can only be in the inactive portion of the genome.   Note that
all never active nodes are ignored by these plots, as their behavior is meaningless.

@@

\section{Conclusion}
\label{sec:conclusion}
\begin{comment}
Even with CGP's relatively simple evolutionary mechanisms, fully understanding
the effects those mechanisms have on search takes concerted effort.  By examining
the effectiveness as well as the actual behavior of traditional CGP and
a handful of variants we have helped advance this understanding.
\end{comment}
